Evaluating the Capabilities of the Lattice Boltzmann Method for non-Newtonian and Free-surface Flow towards Applications in Wellbore Cementing

### 2016-09-10 src/stability-paper.tex
minor updates

Analytical solutions rarely exist for even the simplest non-Newtonian fluid flows because of the complexity that a nonlinear constitutive relationship entails.
It is generally more practical to approximate the flow of non-Newtonian fluids using numerical methods~\cite{syrakos2014performance,owens2002computational,chai2011multiple}.

The lattice Boltzmann method (LBM) is a numerical method for simulating fluid flow that has the advantage (among others) that computing the strain-rate is second-order accurate in space and local to each node~\cite{kruger2010second}.
This means that although an iterative solution is still required to determine the local strain-rate and apparent viscosity, each iterative solution can be done in parallel, by a separate process, as they are independent of each other.
Because hardware architectures have shifted from single, sequential processing systems to parallel processing systems, the local nature of the stress--strain-rate relationship in LBM gives it a distinct advantage for simulating non-Newtonian flow over some other numerical methods.
For example,~\cite{tang2011bingham,chai2011multiple,fallah2012multiple,chen2014simulations,vikhansky2008lattice,wang2008lattice} developed LBM models for simulating yield stress flow.
The LBM model results agreed well when compared to analytical solutions for Bingham plastic Poiseuille flow and values from literature for lid-driven cavity flow, which shows the feasibility of using LBM models for yield-stress fluids.
Examples of LBM models for power-law fluid flow include~\cite{wang2011lattice,wang2015localized,boyd2006second,chai2011multiple}, and blood flows using the K-L, Casson, and Carreau-Yasuda constitutive relationships~\cite{ashrafizaadeh2009comparison}, which have also been successfully verified with benchmarks.

@@

The goal of the present study is to numerically study the implications on accuracy, stability, and efficiency for some of the different strategies for simulating non-Newtonian flow using LBM. The intention of the study is to aid scientists and engineers in understanding which strategy is best suited to their priorities and applications of interest so as to maximize the advantages LBM has in simulating non-Newtonian flow.
Advantages, such as LBM's potential to scale well in parallel, can be much less realized if the collision operator is too computationally expensive, or if numerical instabilities ensue.
A numerical study can help to determine approximate numerical values, domains, and boundary conditions in which one LBM scheme may be more advantageous than another so that LBM may be used in a computationally efficient and stable manner.
In \Fref{sec:LBM}, an overview of the Lattice Boltzmann Method for simulating fluid flow is presented, along with the collision operators, stability enhancements, and constitutive relationships that were investigated.
\Fref{sec:numerical-study} outlines the boundary conditions, material parameters, and LBM collision schemes for each simulation in the numerical study; and is followed by results and discussion.
Finally, concluding remarks are given in \Fref{sec:numerical-study-conclusions}.

@@

\label{sec:picard}
\begin{enumerate}
	\item Start with initial guess for the apparent viscosity, $\mu_k = \mu_0$.
	      Set $\omega_k$ using \eqref{eq:colfreq-non-newtonian} (, or set $s_7 = s_8$ using \eqref{eq:mrt-s7})
	\item \label{step:solve-for-flow} Solve for the strain-rate, $\dot{\gamma}_k$, using \eqref{eq:srtensor-bgk} (, or using \eqref{eq:srtensor-mrt}).
	\item Update the apparent viscosity, $\mu_{k+1} = f(\dot{\gamma}_k)$.
	\item Return to Step \ref{step:solve-for-flow} until convergence is met.
\end{enumerate}
\noindent where $f$ is form of the constitutive relationship that maps the local strain-rate to a local apparent visocisty.
More detail on the functions that are used for $f$ is given in \Fref{sec:constit}.

@@

Due to the discontinuous nature of \eqref{eq:bingham}, the Bingham plastic model is difficult to work with numerically.
Thus, a smooth approximation to \eqref{eq:bingham} formulated by~\citet{papanastasiou1987flows} was used herein, which is defined as:
\begin{equation} \label{eq:papa}
\tau = \tau_y (1 - e^{-m |\dot{\gamma}|}) + \mu_p \dot{\gamma},
\end{equation}
\noindent where $m$ is the stress growth exponent.
The larger the value of $m$, the closer the approximation is to the Bingham plastic model.
Alternatively, the constitutive relationship can be interpreted through the apparent viscosity.
Noting that $\mu_{app} = \frac{\tau}{\dot{\gamma}}$ and rearranging \eqref{eq:papa} results in the following expression for the apparent viscosity:
\begin{equation} \label{eq:bing-mu-app}
  \mu_{app} = f_B(\dot{\gamma}) = \frac{\tau_y}{\dot{\gamma}} (1 - e^{-m |\dot{\gamma}|}) + \mu_p.
\end{equation}

@@

The power-law relationship between stress and strain-rate is given by:
\begin{equation} \label{eq:pl}
\tau = k \dot{\gamma}^n,
\end{equation}
\noindent where $k$ is the flow consistency index and $n$ is the flow behavior index.
When $n = 1$, \eqref{eq:pl} results in a Newtonian constitutive relationship with dynamic viscosity $\mu = k$.
A flow consistency index of $n < 1$ results in shear-thinning behavior, whereas $n > 1$ results in shear-thickening.
As with the Bingham plastic relationship, \eqref{eq:pl} can be modified to determine an apparent viscosity:
\begin{equation} \label{eq:pl-mu-app}
  \mu_{app} = f_{PL}(\dot{\gamma}) = k \dot{\gamma}^{n-1}.
\end{equation}

@@

Before closing the discussion on the Poiseuille flow numerical study, it is worthwhile to consider what precisely is common between the higher-yield stress fluids and the shear-thinning fluids, as they were the cases that exhibited numerical instabilities and were simulated with the least amount of accuracy for some of the collision schemes considered.
One such commonality can be seen when considering the effect of the yield stress, $\tau_y$, and the effect of the flow behavior index, $n$, on the gradient of the theoretical flow velocity.
Taking the partial with respect to $y$ of the analytical solution for the flow velocity of a Bingham plastic, Poiseuille flow (Equation \eqref{eq:analytical-bingham}) results in:
\begin{equation} \label{eq:gradu-bingham}
  \frac{\partial u_x}{\partial y} = \begin{cases}
    0, & 0 \leq |y| \leq y_{\tau}, \\
    \frac{1}{\mu_p} \left( \pgrad y + \tau_y \frac{y}{|y|} \right), & y_{\tau} < |y| \leq h,
  \end{cases}
\end{equation}
and taking the partial with respect to $y$ of the analytical solution for the flow velocity of a power-law fluid, Poiseuille flow (Equation \eqref{eq:analytical-power-law}) results in:
\begin{equation} \label{eq:gradu-power-law}
  \frac{\partial u_x}{\partial y} = \left( -\frac{1}{k} \pgrad \right)^{1/n} \left( -y |y|^{\frac{1}{n} - 1} \right).
\end{equation}
It can be seen from \eqref{eq:gradu-bingham} and \eqref{eq:gradu-power-law} that $\left| \frac{\partial u_x}{\partial y} \right|$ increases with the yield stress for Bingham plastics, and increases with $\frac{1}{n}$ for power-law fluids, which suggests that one of the challenges associated with the accuracy and stability of simulating higher yield-stress fluids and more extreme shear-thinning fluids (e.g. $n \le 0.5$) in Poiseuille-type flow is a result of having a larger magnitude of the gradient of macroscopic flow.
This conclusion is consistent with the discussion in \Fref{sec:bgk} with regards to one of the drawbacks of the BGK collision operator, namely that a sharp spatial gradient in either $\rho$ or $\mvel$ can lead to nonphysical oscillations that can degrade the accuracy and stability of the solution.
However, it does not explain why the MRT collision operator was unable to accurately simulate the case of a power-law fluid, Poiseuille flow, with $n = 0.5$.

@@

Median filtering can be an effective technique for enhancing stability, especially for high Reynold's number flow; however, if the filter is not carefully tuned by properly adjusted the threshold, $\delta$, or number of standard deviations, $n_s$, then the physical intregrity, and consquently accuracy, of the model can be adversely impacted.
In general, the MRT collision operator is much more computationally expensive than its BGK counterpart and is some times even orders of magnitude slower.

To summarize Bingham plastic, lid-driven cavity flow,
    \begin{itemize}
  \item for low Reynold's number flow the BGK collision operator with $m = 10^5$ provides an accurate solution with a short computation time,
    \item for high Reynold's number flow the BGK collision operator requires entropic filtering to remain stable.
    \end{itemize}
To summarize power-law, lid-driven cavity flow,
    \begin{itemize}
      \item for shear-thinning fluids at high Reynold's numbers ($Re \ge 5000$) median filtering is necessary to produce a stable solution,
  \item for shear-thickening, lid-driven cavity flow, there is much less concern for numerical instability.
    \end{itemize}

>>>>

Analytical solutions rarely exist for even the simplest non-Newtonian fluid flows because of the complexity that a nonlinear constitutive relationship entails.
It is generally more practical to approximate the flow of non-Newtonian fluids using numerical methods~\cite{syrakos2014performance,owens2002computational,chai2011multiple}.
The lattice Boltzmann method (LBM) is a numerical method for simulating fluid flow that has the advantage (among others) that computing the strain-rate is second-order accurate in space and local to each node~\cite{kruger2010second}.
This means that although an iterative solution is still required to determine the local strain-rate and apparent viscosity, each iterative solution can be done in parallel, by a separate process, as they are independent of each other.
Because hardware architectures have shifted from single, sequential processing systems to parallel processing systems, the local nature of the stress--strain-rate relationship in LBM gives it a distinct advantage for simulating non-Newtonian flow over some other numerical methods.
For example,~\cite{tang2011bingham,chai2011multiple,fallah2012multiple,chen2014simulations,vikhansky2008lattice,wang2008lattice} developed LBM models for simulating yield stress flow.
The LBM model results agreed well when compared to analytical solutions for Bingham plastic Poiseuille flow and values from literature for lid-driven cavity flow, which shows the feasibility of using LBM models for yield-stress fluids.
Examples of LBM models for power-law fluid flow include~\cite{wang2011lattice,wang2015localized,boyd2006second,chai2011multiple}, and blood flows using the K-L, Casson, and Carreau-Yasuda constitutive relationships~\cite{ashrafizaadeh2009comparison}, which have also been successfully verified with benchmarks.

@@

The goal of the present study is to numerically study the implications on accuracy, stability, and efficiency for some of the different strategies for simulating non-Newtonian flow using LBM. The intention of the study is to aid scientists and engineers in understanding which strategy is best suited to their priorities and applications of interest so as to maximize the advantages LBM has in simulating non-Newtonian flow.
Advantages, such as LBM's potential to scale well in parallel, can be much less realized if the collision operator is too computationally expensive, or if numerical instabilities ensue.
A numerical study can help to determine approximate numerical values, domains, and boundary conditions in which one LBM scheme may be more advantageous than another so that LBM may be used in a computationally efficient and stable manner.
In \Fref{sec:LBM}, an overview of the Lattice Boltzmann Method for simulating fluid flow is presented, along with the collision operators, stability enhancements, and constitutive relationships that were investigated.
\Fref{sec:numerical-study} outlines the boundary conditions, material parameters, and LBM collision schemes for each simulation in the numerical study, and is followed by results and discussion.
Finally, concluding remarks are given in \Fref{sec:numerical-study-conclusions}.

@@

\label{sec:picard}
\begin{enumerate}
  \item Start with initial guess for the apparent viscosity, $\mu_{app}^k = \mu_{app}^0$.
	      Set $\omega^k$ using \eqref{eq:colfreq-non-newtonian} (or set $s_7^k = s_8^k$ using \eqref{eq:mrt-s7})
	\item \label{step:solve-for-flow} Solve for the strain-rate, $\dot{\gamma}^k$, using \eqref{eq:srtensor-bgk} (or using \eqref{eq:srtensor-mrt}).
  \item Update the apparent viscosity, $\mu_{app}^{k+1} = \mu_{app}(\dot{\gamma}^k)$.
	\item Return to Step \ref{step:solve-for-flow} until the apparent viscosity estimate converges.
\end{enumerate}
\noindent where $\mu_{app}(\dot{\gamma}^k)$ is form of the constitutive relationship that maps the local strain-rate to a local apparent visocisty.
More detail on the functions that are used for $\mu_{app}(\dot{\gamma}^k)$ is given in \Fref{sec:constit}.

@@

Due to the discontinuous nature of \eqref{eq:bingham}, the Bingham plastic model is difficult to work with numerically.
Thus, a smooth approximation to \eqref{eq:bingham} formulated by~\citet{papanastasiou1987flows} was used herein, which is defined as:
\begin{equation} \label{eq:papa}
\tau = \tau_y (1 - e^{-m |\dot{\gamma}|}) + \mu_p \dot{\gamma},
\end{equation}
\noindent where $m$ is the stress growth exponent.
The larger the value of $m$, the closer the approximation is to the Bingham plastic model.
Alternatively, the constitutive relationship can be interpreted through the apparent viscosity.
Noting that $\mu_{app} = \frac{\tau}{\dot{\gamma}}$ and rearranging \eqref{eq:papa} results in the following expression for the apparent viscosity:
\begin{equation} \label{eq:bing-mu-app}
  \mu_{app}(\dot{\gamma}) = \frac{\tau_y}{\dot{\gamma}} (1 - e^{-m |\dot{\gamma}|}) + \mu_p.
\end{equation}

@@

The power-law relationship between stress and strain-rate is given by:
\begin{equation} \label{eq:pl}
\tau = k \dot{\gamma}^n,
\end{equation}
\noindent where $k$ is the flow consistency index and $n$ is the flow behavior index.
When $n = 1$, \eqref{eq:pl} results in a Newtonian constitutive relationship with dynamic viscosity $\mu = k$.
A flow consistency index of $n < 1$ results in shear-thinning behavior, whereas $n > 1$ results in shear-thickening.
As with the Bingham plastic relationship, \eqref{eq:pl} can be modified to determine an apparent viscosity:
\begin{equation} \label{eq:pl-mu-app}
  \mu_{app}(\dot{\gamma}) = k \dot{\gamma}^{n-1}.
\end{equation}

@@

Before closing the discussion on the Poiseuille flow numerical study, it is worthwhile to consider what precisely is common between the higher-yield stress fluids and the shear-thinning fluids, as they were the cases that exhibited numerical instabilities and were simulated with the least amount of accuracy for some of the collision schemes considered.
One such commonality can be seen when considering the effect of the yield stress, $\tau_y$, and the effect of the flow behavior index, $n$, on the gradient of the theoretical flow velocity.
Taking the partial derivative with respect to $y$ of the analytical solution for the flow velocity of a Bingham plastic, Poiseuille flow (Equation \eqref{eq:analytical-bingham}) results in:
\begin{equation} \label{eq:gradu-bingham}
  \frac{\partial u_x}{\partial y} = \begin{cases}
    0, & 0 \leq |y| \leq y_{\tau}, \\
    \frac{1}{\mu_p} \left( \pgrad y + \tau_y \frac{y}{|y|} \right), & y_{\tau} < |y| \leq h,
  \end{cases}
\end{equation}
and taking the partial derivative with respect to $y$ of the analytical solution for the flow velocity of a power-law fluid, Poiseuille flow (Equation \eqref{eq:analytical-power-law}) results in:
\begin{equation} \label{eq:gradu-power-law}
  \frac{\partial u_x}{\partial y} = \left( -\frac{1}{k} \pgrad \right)^{1/n} \left( -y |y|^{\frac{1}{n} - 1} \right).
\end{equation}
It can be seen from \eqref{eq:gradu-bingham} and \eqref{eq:gradu-power-law} that $\left| \frac{\partial u_x}{\partial y} \right|$ increases with the yield stress for Bingham plastics, and increases with $\frac{1}{n}$ for power-law fluids, which suggests that one of the challenges associated with the accuracy and stability of simulating higher yield-stress fluids and more extreme shear-thinning fluids (e.g. $n \le 0.5$) in Poiseuille-type flow is a result of having a larger magnitude of the gradient of macroscopic flow.
This conclusion is consistent with the discussion in \Fref{sec:bgk} with regards to one of the drawbacks of the BGK collision operator, namely that a sharp spatial gradient in either $\rho$ or $\mvel$ can lead to nonphysical oscillations that can degrade the accuracy and stability of the solution.
However, it does not explain why the MRT collision operator was unable to accurately simulate the case of a power-law fluid, Poiseuille flow, with $n = 0.5$.

@@

Median filtering can be an effective technique for enhancing stability, especially for high Reynold's number flow. 
However, if the filter is not carefully tuned by properly adjusting the threshold or number of standard deviations then the physical intregrity, and consquently accuracy, of the model can be adversely impacted.
In general, the MRT collision operator is much more computationally expensive than its BGK counterpart and is some times even orders of magnitude slower.
